The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-22 paper.qmd
edit evaluation bit

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.

## The bivariate map
@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is technically* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic did not hide the invalid signal, so it is not performing signal-suppression as we would like. At this point, it might be reasonable to ask, why? Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-bivariate
#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature. This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation. Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the case where the spatial trend could be attributed to noise."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Bivariate Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

## Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006]. The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable. However, @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. If we cannot treat uncertainty as any other variable, what should we treat it as? We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

### It's a variable... it's a metadata... it's uncertainty?
Describing what uncertainty actually is, is surprisingly hard. Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation.

Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*. Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", i.e. we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. Uncertainty is a by-product of inference.

@@

This interpretation of uncertainty indicates that the uncertainty on a statistic is not of value in of itself. Uncertainty is metadata about our statistic that is required for valid inference. This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. This aspect of uncertainty visualization makes it a uniquely difficult problem. This is something frequently mentioned

@@

This means effective uncertainty visualisation should be leveraging integrability. That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020]. When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
#| label: fig-bivariate2
#| layout-ncol: 3
#| layout-valign: "bottom"

#fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature. This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to colour value. Even though colour value is intergrable with colour hue, and colour value is naturally accociated with uncertainty, the spatial trend is still visible in the map."

# Bivariate Map
# Make bivariate palette
# Function to devalue by a certain amount
colsupress2 <- function(basecols, hue=1, sat=1, val=1) {
    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
}

# recurvisely decrease value
v_val = 1.1
bivariatepal2 <- c(basecols,
                   colsupress2(basecols, val=v_val), 
                   colsupress2(colsupress2(basecols, val=v_val), val=v_val),
                   colsupress2(colsupress2(colsupress2(basecols, val=v_val), val=v_val), val=v_val)
                   )

# establish levels of palette
names(bivariatepal2) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# Bivariate maps 2
p2a1 <- my_map_data |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
  scale_fill_manual(values = bivariatepal2) +
  theme_void() + 
  theme(legend.position = "none")

p2b2 <- p2a1 %+% filter(my_map_data, variance_class=="highvar")

p2a1
p2b2
#show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-transparency
#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and variance. This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to transparency."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Transparency Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

# ALTERNATIVE TRANSPARENCY INSTEAD OF COLOUR VALUE
# get colour values from transparency transformation
transpal <- lapply(basecols, scales::alpha, alpha=c(0.95, 0.65, 0.35, 0.05))

transpal <- as.vector(matrix(unlist(transpal), ncol = 4, byrow = TRUE))

names(transpal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

# Transparency maps
pta <- my_map_data |>
  filter(variance_class=="lowvar") %>%
  ggplot() +
  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
  scale_fill_manual(values = transpal) +
  theme_void() + 
  theme(legend.position = "none")

ptb <- pta %+% filter(my_map_data, variance_class=="highvar")

pta
ptb
show_pal2(colours = transpal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)

@fig-transparency is an example of a variations of @fig-bivariate where uncertainty is mapped to transparency, and temperature is mapped to color hue to leverage these visualisation concepts. This method achieves signal suppression quite well. The spatial trend is clearly visible in the low variance case and that trend it becomes much harder to identify in the high variance case. While this was an effective approach for this graphic, relying on integrability may not give us the amount of control we want over our signal-supresison. Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals. Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].

There is another reason @fig-transparency is better at signal-suppression than @fig-bivariate, and it may not be due to integrability. Colour value has a second desirable quality for signal-supression, which is that the colours become harder to distinguish as the value decreases. This means high uncertainty values are harder to differentiate than low uncertainty values. This implicit feature of colour value can generalised to other aesthetics by transforming the visual feature space ourselves.

# Combining uncertainty and signal in a transformed space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.

@@

@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to color the counties. The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. This means the VSUP has successfully suppressed the spatial trend in the data. However the spatial trend may not be the only signal of concern in our graphic. Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met. Are all the justified signals reinforced, while all the unjustified signals are suppressed? Is a graphic that performs perfect signal-suppression even possible?

@@

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another. There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals. If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.

@@

Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA. However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation.

Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights. This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006], however achieving it would largely depend on the methods use to perform signal suppression. By manually combining the values in @fig-vsup, we violated this requirement, so the VSUP is not versatile enough to act as an uncertainty visualisation for EDA.

### The limitations of explicitly visualising uncertainty and signal
The lack of versatility of the VSUP is easy to see with a simple example. Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color. We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". To answer this question we do a series of t-tests on these estimates.

All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other. The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. We cannot perform signal-suppression on both these signals simultaneously.

This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together. Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly. That is, at certain saturation values the colors in a bivariate map are imperceptibly different to the human brain and appear as though they are mapped to the same value. At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain.  Which hypothesis are suppressed and which are not largely depends on the method used to combining colors in the palette [@Kay2019]. The VSUP here used a tree based method as that is what was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.

If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. However, If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.

# Implicitly Combining Uncertainty and Signal
There is technically a stage of our analysis where the estimate and variance are not separate, when we only have a sample. Rather than trying to figure out how to combine signal and uncertainty into a single color, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.

## Pixel map
@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's temperature sampling distribution. The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.

@@

## Show me the data
The pixel map is not the best uncertainty visualisation, but it is trying to *imitate* the best uncertainty visualisation. The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.  As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". It's not a definition that rolls off the tongue, but we can work with it.

@@

What this means is that we have two types of uncertainty, but one is more "processed" than the other. Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic. The disconnect between these two expressions of uncertainty creates a huge problem for authors trying to visualise it. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect. The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty. There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify. Some authors don't believe that it is even possible to quantify all the assumption violations. @Spiegelhalter2017 mentioned that combining the uncertainties that appear at each stage of an analysis into a single "uncertainty" value is near impossible.

@@

This does not mean that visualizing raw data instead of implementing sampling techniques will always prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. While it is not always possible, it should always be considered as an effective uncertainty visualisation when the raw data is available.

# Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", are lacking if we want any generalizable results. However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].

Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consier how we to best evaluate visualisations on their ability to perform signal-suppression.

@@

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. Science communication should be primarily concerned with accuracy.

@@

Another metric that is similar to trust is the participants confidence in their decision or extracted value. Confidence has many of the same issues as trust, but it has an additional confounding factor. In non-uncertainty visualisation evaluation experiments, "confidence" is used as a proxy for the clarity of the visualisation. Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. Uncertainty visualisations conflate these two measures when they ask about confidence.

@@

The first method is what we call the "vague question" approach. The authors of these studies will ask the participants a question that implies that they should use  a use of uncertainty, but compare the readers answer to a very specific ground truth. This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right". The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution. Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

>>>>

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This is a reasonable approach. This is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. However, it is unclear if this approach is will for our purposes.

## The bivariate map
@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional color palette. In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist). The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like. At this point, it might be reasonable to ask, why? Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-bivariate
#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature. This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation. Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the high variance case."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Bivariate Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

## Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006]. While this is part of the problem in uncertainty visualisation, it is not the complete picture. @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. If we cannot treat uncertainty the same as we would any other variable, how should we treat it as? We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

### It's a variable... it's a metadata... it's uncertainty?
Describing what uncertainty actually is, is surprisingly hard. Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*.

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.

@@

This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself. Uncertainty is metadata about our statistic that is required for valid inference. This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. This aspect of uncertainty visualization makes it a uniquely difficult problem.

@@

This means effective uncertainty visualisation should be leveraging integrability. That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020]. When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012]. Unfortunately relying on integrability may not give us the amount of control we want over our signal-supresison. Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals. Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].

There is another benefit to mapping uncertainty to stauration that is not directly related to integrability. Colour value has a second desirable quality as the colours become harder to distinguish as the saturation decreases. This means high uncertainty saturations are harder to differentiate than low uncertainty values. We can leverage this implicit feature of colour value by transforming the visual feature space ourselves.

# Combining uncertainty and signal in a transformed space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. As a matter of fact, some uncertainty visualisation authors already have.

@@

@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to color the counties. The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. This means the VSUP has successfully suppressed the spatial trend in the data. However, the spatial trend may not be the only signal of concern in our graphic. Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met. Are all the justified signals reinforced, while all the unjustified signals are suppressed? Is a graphic that performs perfect signal-suppression even possible?

@@

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another. There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals. If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. Unfortunately it is unclear how reliably this would work.

@@

Some authors recognize inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.

### The limitations of explicitly visualising uncertainty and signal
The lack of versatility of the VSUP is easy to see with a simple example.

Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color. We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". To answer this question we do a series of t-tests on these estimates. All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other. The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. We cannot perform signal-suppression on both these signals simultaneously.

This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together. Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly. That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value. At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain.

If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. This issue has already been raised in the literature. Which hypothesis are suppressed and which are not largely depends on the method used to combining colors in the palette [@Kay2019]. The VSUP in @fig-vsup used the tree based method that was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.

Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights. This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006]. If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.

# Implicitly Combining Uncertainty and Signal
Rather than trying to figure out how to combine signal and uncertainty into a single color, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.

## Pixel map
@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's (i.e. the county's) average temperature sampling distribution. The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.

@@

## Show me the data
The pixel map is not the best uncertainty visualisation, but it does share a lot in common with the best uncertainty visualisation. The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.  As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". It's not a definition that rolls off the tongue, but we can work with it.

@@

What this means is that we have two types of uncertainty, but one is more "processed" than the other. Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data). The disconnect between quantifiable and unqualitifable uncertainty creates a huge problem for authors trying to visualize it. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect. The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty. There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify. Some authors don't believe that it is even possible to quantify all the assumption violations [@Spiegelhalter2017].

@@

This does not mean that visualizing raw data will always prevent insignificant signal from getting through. @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked.

While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. In order to uncerstand *why* pixel maps seem to be successful at signal suppression at first glace would likely require a more thorough discussion than what what we can offer here.

# Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results. However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].

Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consider how to evaluate visualisations on their ability to perform signal-suppression.

@@

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018]. Science communication should be primarily concerned with accuracy.

@@

Another metric that is similar to trust is the participants confidence in their decision or extracted value. Confidence has many of the same issues as trust, but it has an additional confounding factor. In non-uncertainty visualisation evaluation experiments, "confidence" is used as a proxy for the clarity of the visualisation. Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. Uncertainty visualisations conflate these two measures.

@@

The first method is what we call the "vague question" approach. The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. The authors then go on to compare the readers answer to a very specific ground truth. This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right". The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution. Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. These studies frequently expect participants to give deterministic answers for probabilistic questions. For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and aksed them "how likely is candidate B to win the election?". Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. Interestingly it is unclear how the participants were supposed to answer this question. @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage) but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
